---
title: "Practice 5: XAI - Partial Dependence Plots"
author: "Andres Alvarruiz Lopez, Alejandro Legleye Enguix, Ferran Soler Martinez"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry: margin=2.2cm
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 4.6
)

set.seed(123)

required_packages <- c(
  "readr",
  "dplyr",
  "ggplot2",
  "randomForest",
  "gridExtra",
  "viridis"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(readr)
library(dplyr)
library(ggplot2)
library(randomForest)
library(gridExtra)
library(viridis)

theme_set(theme_minimal(base_size = 10))

fmt_pct <- function(x) paste0(round(100 * x, 1), "%")
fmt_k <- function(x) paste0(round(x / 1000), "k")

pdp_1d <- function(model, data, variable, grid) {
  values <- vapply(grid, function(value) {
    modified_data <- data
    modified_data[[variable]] <- value
    mean(predict(model, newdata = modified_data))
  }, numeric(1))

  tibble::tibble(
    variable = variable,
    value = grid,
    prediction = values
  )
}

pdp_2d <- function(model, data, variable_x, variable_y, grid_x, grid_y) {
  grid <- expand.grid(
    value_x = grid_x,
    value_y = grid_y,
    KEEP.OUT.ATTRS = FALSE
  )

  grid$prediction <- vapply(seq_len(nrow(grid)), function(i) {
    modified_data <- data
    modified_data[[variable_x]] <- grid$value_x[i]
    modified_data[[variable_y]] <- grid$value_y[i]
    mean(predict(model, newdata = modified_data))
  }, numeric(1))

  tibble::as_tibble(grid)
}

regular_grid <- function(x, n = 50, probs = c(0.01, 0.99)) {
  limits <- as.numeric(stats::quantile(x, probs = probs, na.rm = TRUE))
  seq(limits[1], limits[2], length.out = n)
}
```

# Introduction

The objective of this practice is to apply model-agnostic explainability
methods. Specifically, Partial Dependence Plots (PDP) are used to analyze how
the average prediction of a random forest changes when one variable is forced to
take different values.

A PDP does not show a causal relationship. It shows the relationship learned by
the model while keeping the remaining variables distributed as observed in the
data. For this reason, PDPs are useful for interpreting nonlinear models, but
they should be read as explanations of the model rather than as direct laws of
the real phenomenon.

```{r load-data}
bike_raw <- read_csv("day.csv", show_col_types = FALSE)
house_raw <- read_csv("kc_house_data.csv", show_col_types = FALSE)
```

# 1D PDP: Bike Rental Prediction

```{r bike-model}
bike_data <- bike_raw %>%
  mutate(
    season = factor(season),
    yr = factor(yr),
    mnth = factor(mnth),
    holiday = factor(holiday),
    weekday = factor(weekday),
    workingday = factor(workingday),
    weathersit = factor(weathersit)
  ) %>%
  select(-dteday, -casual, -registered)

bike_x <- bike_data %>% select(-cnt)
bike_y <- bike_data$cnt

bike_rf <- randomForest(
  x = bike_x,
  y = bike_y,
  ntree = 300,
  mtry = 4,
  importance = TRUE
)

bike_oob_r2 <- tail(bike_rf$rsq, 1)

bike_pdp <- bind_rows(
  pdp_1d(
    bike_rf,
    bike_x,
    "instant",
    seq(min(bike_x$instant), max(bike_x$instant), length.out = 60)
  ),
  pdp_1d(bike_rf, bike_x, "temp", regular_grid(bike_x$temp, n = 60)),
  pdp_1d(bike_rf, bike_x, "hum", regular_grid(bike_x$hum, n = 60)),
  pdp_1d(
    bike_rf,
    bike_x,
    "windspeed",
    regular_grid(bike_x$windspeed, n = 60)
  )
) %>%
  mutate(
    variable_label = recode(
      variable,
      instant = "Days since 2011",
      temp = "Normalized temperature",
      hum = "Normalized humidity",
      windspeed = "Normalized wind speed"
    )
  )

stopifnot(!anyNA(bike_pdp$prediction))
```

A regression random forest is fitted to predict `cnt`, the total number of bikes
rented per day. The variables `casual` and `registered` are removed because they
are direct components of `cnt` and would introduce data leakage. The approximate
OOB explained variance of the model is `r fmt_pct(bike_oob_r2)`.

```{r bike-pdp-plot}
ggplot(bike_pdp, aes(x = value, y = prediction)) +
  geom_line(color = "#1f6f8b", linewidth = 0.9) +
  facet_wrap(~ variable_label, scales = "free_x", ncol = 2) +
  labs(
    title = "1D PDP for the bike rental model",
    x = "Fixed value of the variable",
    y = "Average predicted cnt"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

**Interpretation.** The number of days since 2011 has an increasing effect: the
model learns a positive time trend, consistent with higher demand in the second
part of the observed period. Temperature also has a clearly positive effect up
to medium-high values; with more pleasant temperatures, the model predicts more
bike rentals.

Humidity shows the opposite pattern: as humidity increases, especially at high
levels, the average predicted number of bikes decreases. Wind speed also reduces
predicted demand when it reaches high values, although its effect is weaker and
more irregular than the effects of temperature or humidity.

# 2D PDP: Humidity and Temperature

```{r bike-2d-data}
bike_pdp_sample <- bike_x %>%
  slice_sample(n = min(500, nrow(bike_x)))

temp_grid <- regular_grid(bike_x$temp, n = 30)
hum_grid <- regular_grid(bike_x$hum, n = 30)

bike_pdp_2d <- pdp_2d(
  bike_rf,
  bike_pdp_sample,
  "temp",
  "hum",
  temp_grid,
  hum_grid
)

stopifnot(!anyNA(bike_pdp_2d$prediction))

temp_step <- diff(range(temp_grid)) / (length(temp_grid) - 1)
hum_step <- diff(range(hum_grid)) / (length(hum_grid) - 1)

plot_2d <- ggplot(bike_pdp_2d, aes(x = value_x, y = value_y, fill = prediction)) +
  geom_tile(width = temp_step, height = hum_step) +
  scale_fill_viridis_c(option = "C", name = "Average\nprediction") +
  labs(
    x = "Normalized temperature",
    y = "Normalized humidity",
    title = "2D PDP: temperature and humidity"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

temp_density <- ggplot(bike_pdp_sample, aes(x = temp)) +
  geom_density(fill = "#1f6f8b", alpha = 0.65, color = NA) +
  labs(x = NULL, y = "Density") +
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 8),
    panel.grid.minor = element_blank()
  )

hum_density <- ggplot(bike_pdp_sample, aes(x = hum)) +
  geom_density(fill = "#b45f35", alpha = 0.65, color = NA) +
  coord_flip() +
  labs(x = NULL, y = "Density") +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_text(size = 8),
    panel.grid.minor = element_blank()
  )

empty_plot <- ggplot() + theme_void()
```

For the two-dimensional PDP, a random sample of 500 observations is used, as
requested in the assignment. The average prediction is computed for each
combination of temperature and humidity values.

```{r bike-2d-plot, fig.width=7.4, fig.height=6.8}
grid.arrange(
  temp_density,
  empty_plot,
  plot_2d,
  hum_density,
  layout_matrix = rbind(c(1, 2), c(3, 4)),
  widths = c(4, 1.2),
  heights = c(1.1, 4)
)
```

**Interpretation.** The highest prediction area appears with medium or high
temperature and low or moderate humidity. When humidity is high, the predicted
number of rentals decreases even if the temperature is favorable. This indicates
that the model learns not only individual effects, but also an intuitive
interaction: warm days favor bike rental, but if they are accompanied by high
humidity, the expected demand decreases.

The marginal densities help identify where real observations are located. Areas
of the map with low density should be interpreted more carefully, because the
model is extrapolating more than in frequent regions of the data.

# PDP: House Price Prediction

```{r house-model}
house_data <- house_raw %>%
  select(
    price,
    bedrooms,
    bathrooms,
    sqft_living,
    sqft_lot,
    floors,
    yr_built
  ) %>%
  filter(
    bedrooms >= 1,
    bedrooms <= 8,
    bathrooms > 0,
    floors > 0,
    sqft_living > 0,
    sqft_lot > 0
  )

house_x <- house_data %>% select(-price)
house_y <- house_data$price

house_rf <- randomForest(
  x = house_x,
  y = house_y,
  ntree = 300,
  mtry = 3,
  importance = TRUE
)

house_oob_r2 <- tail(house_rf$rsq, 1)

house_pdp_sample <- house_x %>%
  slice_sample(n = min(2000, nrow(house_x)))

bedroom_grid <- sort(unique(house_x$bedrooms))
bathroom_grid <- sort(unique(house_x$bathrooms))
bathroom_grid <- bathroom_grid[
  bathroom_grid >= as.numeric(quantile(house_x$bathrooms, 0.01)) &
    bathroom_grid <= as.numeric(quantile(house_x$bathrooms, 0.99))
]
sqft_living_grid <- regular_grid(house_x$sqft_living, n = 60)
floor_grid <- sort(unique(house_x$floors))

house_pdp <- bind_rows(
  pdp_1d(house_rf, house_pdp_sample, "bedrooms", bedroom_grid),
  pdp_1d(house_rf, house_pdp_sample, "bathrooms", bathroom_grid),
  pdp_1d(house_rf, house_pdp_sample, "sqft_living", sqft_living_grid),
  pdp_1d(house_rf, house_pdp_sample, "floors", floor_grid)
) %>%
  mutate(
    variable_label = recode(
      variable,
      bedrooms = "Bedrooms",
      bathrooms = "Bathrooms",
      sqft_living = "Living area",
      floors = "Floors"
    )
  )

stopifnot(!anyNA(house_pdp$prediction))
```

For house price prediction, a random forest is trained with the variables
specified in the assignment: `bedrooms`, `bathrooms`, `sqft_living`, `sqft_lot`,
`floors`, and `yr_built`. The approximate OOB explained variance of the model is
`r fmt_pct(house_oob_r2)`. To compute the PDPs, a sample of 2000 observations is
used, reducing the computational cost without changing the logic of the method.

```{r house-pdp-plot}
ggplot(house_pdp, aes(x = value, y = prediction)) +
  geom_line(color = "#31572c", linewidth = 0.9) +
  geom_point(color = "#31572c", size = 1.4) +
  facet_wrap(~ variable_label, scales = "free_x", ncol = 2) +
  scale_y_continuous(labels = fmt_k) +
  labs(
    title = "1D PDP for the house price model",
    x = "Fixed value of the variable",
    y = "Average predicted price"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

**Interpretation.** The variable with the clearest effect is `sqft_living`: as
living area increases, the predicted price also increases. This pattern is
expected because interior size is one of the most direct signals of a house's
value.

Bathrooms also increase the predicted price, although with plateau regions. This
suggests that the model values the transition from few bathrooms to a medium
number more than adding bathrooms when the house already has several. Bedrooms
show a less linear effect: more bedrooms do not always imply a higher price if
they are not accompanied by more living area or other characteristics. Floors
have a moderate and irregular effect; the model seems to associate them with
some increase in price, but not with a relationship as strong as living area.

# Conclusions

PDPs make it possible to visually summarize the relationships learned by random
forest models. In the bike rental case, temperature and the time trend increase
the prediction, while high humidity and wind speed reduce it. In the house price
case, living area dominates the explanation of price, followed by bathrooms;
bedrooms and floors have more context-dependent effects.

The main precaution is that PDPs average over all observations. If variables are
highly correlated or some regions have few data points, the interpretation must
be made with caution.
